Report: Complex Support Vector Detector for Large MIMO System

### 2015-10-02 Report Aug 5 2015/Report Aug 2015.tex
Report Aug 5 revised
Introduction

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

%\hfill September 17, 2014
  One of the biggest challenges the researchers and industry practitioners are facing in wireless communication area is how to bridge the sharp gap between increasing demand of high speed communication of rich multimedia information with high level Quality of Service (QoS) and the limited radio frequency spectrum over a complex space-time varying environment. As the most promising technology for solving this problem, Multiple Input Multiple Output (MIMO) technology has been of immense research interest over the last several tens of years and become mature, which is incorporated into the emerging wireless broadband standard like 802.11ac\cite{IEEE802.11ac} long-term evolution (LTE)\cite{3GLTE}.  The core idea of MIMO system is to use multiple antennas at both transmitting and receiving end, so that multiplexing gain (multiple parallel spatial data pipelines that can improve bandwidth efficiency) and diversity gain (better reliability of communication link) is obtained by employing spatial domain. Large MIMO (also called Massive MIMO) is an upgrade version of conventional MIMO system, it equips unprecedented hundreds of low power low price antennas at base station (BS), serving several tens of terminals simultaneously, It can achieve full potential of conventional MIMO system while providing additional power efficiency as well as system robustness\cite{rusek2013scaling}\cite{larsson2014massive}.

  The price paid for large MIMO system is the increasing complexities for signal processing at both transmitting and receiving end. Uplink Detector is one of the key components in large MIMO system. With orders magnitude more antennas equipped at BS, benefits and challenges coexist in designing of detection algorithms for large MIMO uplink, on the one hand, large number of receive antennas provide potential of large diversity gain, on the other hand, complexity of the algorithm becomes extremely crucial to make system practical.

The optimal maximum likelihood detector (MLD) for MIMO system requires the complexity increase exponentially with number of transmitted antennas with a factor of the size of constellation, which is prohibitive in practical implementations. 
Sphere Decoder (SD)\cite{damen2003maximum} is the most prominent algorithm that utilizes lattice structure of MIMO system. 
Its variant fixed complexity sphere decoder (FCSD)\cite{barbero2008fixing} make it possible to achieve near optimal performance with a fixed complexity under different signal to noise ratio (SNR). However, all the algorithms that based on lattice structure have the same shortage - their complexities increases exponentially with a factor of the size of symbol constellation. Therefore, they are prohibitive when it comes to a high order modulation scheme, for example in IEEE 802.11ac standard\cite{IEEE802.11ac}, the modulation scheme is 256QAM.

Linear detectors (LD) like minimum mean square error (MMSE) and zero forcing (ZF) along with their sequential interference cancellation with optimized ordering (OSIC) counterparts\cite{wolniansky1998v}\cite{foschini1999simplified}\cite{benesty2003fast}, which have good performance for low loading factor in massive MIMO system (the number of receive antennas is much larger than the number of transmit antennas)\cite{hoydis2013massive}. In the last several years, a set of detection algorithms are proposed with complexities that is comparable with LD-OSIC and suboptimal performance can be achieved.  The local search algorithm, such as likelihood ascend searching (LAS)\cite{vardhan2008low}\cite{li2010multiple}, an theoretical analysis of upper bound of bit error rate (BER) and lower bound of on asymptotic multiuser efficiency for the LAS detector was presented\cite{sunfamily}. Layered Tabu search algorithm presented in\cite{srinidhi2011layered} is superior to the LAS algorithms because it can move away to new searching area to avoid local minimal. Message passing detectors based on belief propagation (BF) and Gaussian Approximation (GA) \cite{goldberger2011mimo}\cite{som2011low}\cite{narasimhan2014channel}\cite{som2010improved}. Markov Chain Monte Carlo (MCMC) algorithm\cite{datta2013novel}  and Lattice Reduction (LR) aided detectors\cite{zhou2013element}.

Firmly grounded in framework of statistical learning theory, Support Vector Machine (SVM) has become a powerful tool to solve real world supervised learning problems such as classification, regression and prediction. SVM method is a nonlinear generalization of Generalized Portrait algorithm developed by Vapnik in 1960s\cite{vapnik1963pattern}\cite{vapnik1964note}, which can give good generalization performance to unseen data\cite{scholkopf2002learning}.

Research and industry interest of SVM boosted since 1990s, promoted by related works of Vapnik and co-workers at AT\& T Bell laboratory\cite{boser1992training}\cite{guyon1993automatic}\cite{vapnik2013nature}\cite{cortes1995support}\cite{scholkopf1996incorporating}\cite{vapnik1996support}.
Moreover, the kernel based methods\cite{scholkopf2002learning} carries out nonlinear learning task by mapping input data sets into high dimensional feature space, then replacing inner product of feature mappings by computational inexpensive kernel functions discarding the actual structure of the feature space. This rationale is supported mathematically by Reproducing Kernel Hilbert Space (RKHS). 
Based on the same regularized risk function principle, $\epsilon$-Support Vector Regression ($\epsilon$-SVR) was developed\cite{vapnik2013nature}\cite{smola2004tutorial}.

Like SVM, $\epsilon$-SVR solving original optimization problem by transforming it into Lagrange dual optimization problem, which can be solved by Quadratic Programming (QP), Sequential Minimal Optimization (SMO) algorithm was proposed as a fast algorithm to solve this QP problem by decompose the it into sub QP problems and solve them analytically\cite{platt1999fast}, therefore, the computational intensive numerical method can be avoided. A more general method is decomposition solver, which refers to a set of algorithms that separate the optimization variables (Lagrange multipliers) into two sets W and N, W is the work set and N contains the remaining optimization variables. In each iteration, only the optimization variables in work set is optimized while keeping other variables fixed. SMO algorithm is an extreme case of decomposition solver. An important issue of decomposition solver is the choice of work set, one strategy is to choose Karush-Kuhn-Tucker (KKT) condition violators, and final converge can be guaranteed\cite{osuna1997improved}. SMO algorithm restricts the size of work set to 2, because of linear constraint in dual problem that inducted by offset. In\cite{steinwart2011training}, a method to train SVM without offset was proposed, with the comparable performance to the SVM with offset. The authors work demonstrates that with the combination of two single optimization variable work set selection strategies which requires searching time $O(n)$ and update a work set size of two in each iteration, this method can achieve a iteration time as few as that searching over all pairs of optimization variables which requires $O(n^{2})$ searching times.

Until now, although the mathematical foundation of kernel based methods is RKHS which is defined in complex domain, the most of practitioners are dealing with real data set. However in communication and signal processing area, the channel gains, signals, waveforms etc. are all represented in complex form. Recently, a pure complex SVR \& SVM based on complex kernel was proposed in\cite{bouboulis2013complex}, which can process the complex data set purely in complex domain. The simulation of channel realization and equalization in\cite{bouboulis2013complex} demonstrate a better performance as well as reduced complexity comparing to simply split learning task into two real case by real kernel.  
Based on this work, we construct a prototype of a complexity-performance controllable detector for large MIMO based on dual channel complex SVR. The detector can work in two parallel real SVR pipeline which can be solved independently. Moreover, only real part of kernel matrix is needed in both channel. This means a large amount of computation can be reduced.
Based on the discrete time MIMO channel model, In our regression model, this CSVR-detector
is constructed without offset, Therefore, for each real SVR without offset, in principle, only one variable is needed to be updated in each iteration, In our prototype, we propose a sequential single optimization variable searching strategy that find two optimization variable sequentially, which can approximate the optimal double optimization variables searching strategy.

>>>>

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

@@

%\hfill September 17, 2014
  One of the biggest challenges the researchers and industry practitioners are facing in wireless communication area is how to bridge the sharp gap between increasing demand of high speed communication of rich multimedia information with high level Quality of Service (QoS) requirements and the limited radio frequency spectrum over a complex space-time varying environment. The most promising technology for solving this problem, Multiple Input Multiple Output (MIMO) technology has been of immense research interest over the last several tens of years is incorporated into the emerging wireless broadband standard like 802.11ac\cite{IEEE802.11ac} long-term evolution (LTE)\cite{3GLTE}.  The core idea of MIMO system is to use multiple antennas at both transmitting and receiving end, so that multiplexing gain (multiple parallel spatial data pipelines that can improve bandwidth efficiency) and diversity gain (better reliability of communication link) is obtained by exploiting the spatial domain. Large MIMO (also called Massive MIMO) is an upgraded version of conventional MIMO technology employing hundreds of low power low price antennas at base station (BS), that serves several tens of terminals simultaneously. This technology can achieve full potential of conventional MIMO system while providing additional power efficiency as well as system robustness both to unintended man-made interference and intentional jamming.\cite{rusek2013scaling}\cite{larsson2014massive}.

  The price paid for large MIMO system is the increased complexities for signal processing at both transmitting and receiving end. The uplink Detector is one of the key components in large MIMO systems. With orders magnitude more antennas at the BS, benefits and challenges coexist in designing of detection algorithms for the uplink communication of large MIMO systems. On one hand, a large number of receive antennas provide potential of large diversity gains, on the other hand, complexity of the algorithm becomes crucial to make the system practical.

Vertical Bell Laboratories Layered Space-Time (V-BLAST) architecture for MIMO system can achieve high spectrum efficiency by spatial multiplexing (SM), that is, each transmit antenna transmits independent symbol streams. However the optimal maximum likelihood detector (MLD) for V-BLAST systems that perform exhaustive search has a complexity that increases exponentially with number of transmitted antennas, which is prohibitive for practical applications.

As alternatives to MLD, linear detectors (LD) such as zero-forcing (ZF) and minimum mean square error (MMSE) with optimized ordering sequential interference cancellation (ZF-OSIC, MMSE-OSIC) are exploited in V-BLAST architecture\cite{wolniansky1998v}\cite{foschini1999simplified}\cite{benesty2003fast},  however the performance of ZF-OSIC and MMMSE-OSIC are inferior comparing to MLD.

Sphere Decoder (SD)\cite{damen2003maximum} is the most prominent algorithm that utilizes the lattice structure of MIMO systems, which can achieve optimal performance with relatively much lower complexity comparing to MLD. However, SD has two major shortages that make it problematic to be integrated into a practical systems. The first shortage is SD has various complexities under different signal to noise ratios (SNR), while a constant processing data rate is required for hardware. The second shortage is SD's complexity still has a lower bound for complexity that increases exponentially with the number of transmit antennas and the order of modulation scheme\cite{jalden2005complexity}.
The fixed complexity sphere decoder (FCSD)\cite{barbero2008fixing} makes it possible to achieve near optimal performance with a fixed complexity under different value of SNR. The FCSD inherits the principle of list based searching algorithms, which first generate a list of candidate symbol vectors and then the best candidate is chosen as the solution. The other sub optimal detectors belong to this class include Generalized Parallel Interference Cancellation (GPIC)\cite{luo2008generalized} and Selection based MMSE-OSIC(sel-MMSE-OSIC)\cite{radji2009interference}. However, all these list based searching algorithms have the same shortage - their complexities increase exponentially with the number of transmit antennas and the order of modulation scheme\cite{radji2009interference}. Therefore, such algorithms are prohibitive when it comes to a large number of antennas or a high order modulation scheme, for example in IEEE 802.11ac standard\cite{IEEE802.11ac}, the modulation scheme is 256QAM.

%Linear detectors (LD) like minimum mean square error (MMSE) and zero forcing (ZF) along with their sequential interference cancellation with optimized ordering (OSIC) counterparts\cite{wolniansky1998v}, which have good performance for low loading factor in massive MIMO system (the number of receive antennas is much larger than the number of transmit antennas)\cite{hoydis2013massive}. 
Besides the above detection algorithms designed for conventional MIMO systems, in the last several years, a set of detection algorithms have been proposed for large MIMO systems with complexities that are comparable with MMSE detector and near-optimal performance. such algorithms include likelihood ascend searching (LAS) algorithms\cite{vardhan2008low}\cite{li2010multiple}, %An theoretical analysis of upper bound of bit error rate (BER) and lower bound of on asymptotic multiuser efficiency for the LAS detector was also presented\cite{sunfamily}.
Tabu search based algorithms which have superior performance compared to LAS detectors because local minima can be avoided (e.g. Layered Tabu search (LTS)\cite{srinidhi2011layered}, Random Restart Reactive Tabu search (R3TS)\cite{datta2010random}), Message passing technique based algorithms (e.g. Belief propagation (BP) detectors based on graphic model and Gaussian Approximation (GA)\cite{som2011low}\cite{som2010improved}\cite{narasimhan2014channel}\cite{goldberger2011mimo}), Probabilistic Data Association based algorithms \cite{mohammed2009low}, Monte Carlo sampling based algorithms (e.g. Markov Chain Monte Carlo (MCMC) algorithm\cite{datta2013novel}) and Lattice Reduction (LR) aided algorithms\cite{zhou2013element}.

Firmly grounded in framework of statistical learning theory, the Support Vector Machine (SVM) technique has become a powerful tool to solve real world supervised learning problems such as classification, regression and prediction. the SVM method is a nonlinear generalization of Generalized Portrait algorithm developed by Vapnik in 1960s\cite{vapnik1963pattern}\cite{vapnik1964note}, which can provide good generalization performance\cite{scholkopf2002learning}.

Interest in SVM boosted since 1990s, promoted by the works of Vapnik and co-workers at AT\& T Bell laboratory\cite{boser1992training}\cite{guyon1993automatic}\cite{vapnik2013nature}\cite{cortes1995support}\cite{scholkopf1996incorporating}\cite{vapnik1996support}.
Moreover, the kernel based methods\cite{scholkopf2002learning} solve nonlinear learning tasks by mapping input data sets into high dimensional feature spaces, and replacing inner products of feature mappings by computational inexpensive kernel functions discarding the actual structure of the feature space. This rationale is supported mathematically by the notion of Reproducing Kernel Hilbert Space (RKHS). 
Based on the same regularized risk function principle, $\epsilon$-Support Vector Regression ($\epsilon$-SVR) was developed\cite{vapnik2013nature}\cite{smola2004tutorial}.

Similar to SVM, the $\epsilon$-SVR solves an original optimization problem by transforming it into a Lagrange dual optimization problem, which can be solved by Quadratic Programming (QP). Sequential Minimal Optimization (SMO) algorithm was proposed as a fast algorithm to solve this QP problem by decomposing the it into sub QP problems and solving them analytically\cite{platt1999fast}. Therefore, the computational intensive numerical method can be avoided. A more general method is decomposition solver, which refers to a set of algorithms that separate the optimization variables (Lagrange multipliers) into two sets W and N, W is the work set and N contains the remaining optimization variables. In each iteration, only the optimization variables in the work set is optimized while keeping other variables fixed. The SMO algorithm is an extreme case of decomposition solver. An important issue of decomposition solver is the choice of the work set. One strategy is to choose Karush-Kuhn-Tucker (KKT) condition violators, ensuring final converge\cite{osuna1997improved}. Because of the linear constraint inducted by offset, the SMO algorithm restricts the size of the work set to 2. In\cite{steinwart2011training}, a method to train SVM without offset was proposed, with the comparable performance to the SVM with offset. A set of sequential single variable work set selection strategies, which require $O(n)$ searching time are proposed. The optimal double variable work set selection strategy, which performs exhaustively searching, however, requires $O(n^{2})$ searching time. The authors demonstrate that with the combination of two proposed single variable work set selection strategies, convergence can be achieved by a iteration time that is as few as optimal double variable work set selection strategy.

The mathematical foundation of kernel based methods is RKHS which is defined in complex domain, however most of the practitioners are dealing with real data sets. In communication and signal processing area, the channel gains, signals, waveforms etc. are all represented in complex form. Recently, a pure complex SVR \& SVM based on complex kernel was proposed in\cite{bouboulis2013complex}, which can deal with the complex data set purely in complex domain. The results in\cite{bouboulis2013complex} demonstrate better performances as well as reduced complexity comparing to simply split learning task into two real case by real kernel.  
Based on this work, we derive of a complexity-performance controllable detector for large MIMO systems based on a dual channel complex SVR (CSVR). The detector can work in two parallel real SVR channels which can be solved independently. Moreover, only the real part of kernel matrix is needed in both channels. This means a large amount of computation can be reduced.
Based on the discrete time MIMO channel model, in our regression model, this CSVR-detector
is constructed without offset, Therefore, for each real SVR without offset, in principle, only one variable is needed to be updated in each iteration, In our scheme, a sequential single variable selection strategy is proposed. By this strategy, two variables can be updated at each iteration, with much smaller searching time.
